Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-02 tagless/tagless-final.tex
more small scale edits

Because |RCN| has the signature Symantics, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the following example.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun self -> self)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
Section~\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ();;
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end;;
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with Reynolds (cite) (albeit in a typed setting). The interpreter RCV is
useful for achieving the CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

In this section however we show a more general approach to CBV CPS:
a CPS transformer. We take any implementation of |Symantics| (interpreter,
compiler, partial evaluator) and make it evaluate the CPS language
instead. We can view this transformation of interpreters as a
transformation on the object language, from direct to
continuation-passing style.

The transformation is canonical, as the following (abbreviated) code
shows
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
  let if_ ec et ef = S.lam (fun k ->
    S.app ec (S.lam (fun vc ->
    S.if_ vc (fun () -> S.app (et ()) k) (fun () -> S.app (ef ())
    k))))
  ...
  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
    f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
  let fix f  = S.fix (fun self -> (f self))
end
\end{code}
The code literally establishes a mapping from CPS interpretations to
(direct) interpretations, that latter performed by 
the base interpreter |S|.

A reader may notice that the above module does not define the type
equation for |repr| and so it does not have the signature Symantics.
The reason for that is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has the type
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we need a 
type case-discriminating
function similar to the type function |repr_pe| of
Section~\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom, in that case) type variable.

To save space, we show a simplified and less principal approach that
lets us use the module |CPST| as it is. Because the module does not
have the signature Symantics, we cannot apply the |EX| functor to it.
But we can write the tests nevertheless:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. The sample terms are \emph{exactly}
as we used before. We just copied them over. That copying step is the
price we pay for the simplified treatment. We shall encounter this
copying when discussing the self-interpreter in the next section. We
will see then that the copying step has a profound meaning of
`plugging the hole', which is one of the faces of polymorphism.

@@

which shows the naive CPS transform of the original code. The result is
quite unoptimal with several apparent $\beta$-redexes. We can reduce
them by instantiating |CPST| with the |P| evaluator instead. Our
running test gives then
\begin{code}
  P.repr = {P.st = Some <fun>; P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We can easily modify the CBV CPS transform to pass a piece of state
along with the continuation. This lets us support mutable state. We
can also use reference variables of the metalanguage to emulate
reference variables in the object language.
\oleg{Should we have an example in the code?}

@@

Taha et al (cite different paper) define a self-interpreter si to be an
object-language term such that, for any object-language term e,

@@

where |"e"| is an encoding of |e|.  Under the co-approach, we want to encode
each object-language term as a function from a record of cogen methods
to a generated rep.  (Thanks to currying, the object language need not
actually support records.)  In the case of a self-interpreter, that
record is simply si!  So we should define a self-interpreter si to be an
object-language record-term such that, for any object-language term e,

@@

The self-interpreter then is just like the interpreter
|R| in Section 2, except written at the object rather than meta level.

Note on terms and special forms. In our class Symantics, |add|
is not a term -- because it does not have the type repr t. We could
have introduced |add| to be of the type |repr (Int->Int->Int)| -- but we
didn't. So, |add| is a special form. It is to be applied in
meta-language (using the Haskell application -- white space) rather
than in the object language (app).

Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
RR is not an identity: it is an interpreter that encapsulates another
interpreter. The RR interpreter can be made self if we treat RR as a
special form and interpret it always with itself (similar to let and
hole below).

Self-interpretation in our framework is
a bit tricky due to a need for polymorphism.  An encoded term should
be polymorphic in the interpreter type-constructor "r".  Further, the
functions "lam", "app", etc. above should be polymorphic in the object
types "a" and "b".  Thus the object language seems to require rank-2
polymorphism; and so type annotations will be inevitable.

To neatly bypass the need for rank-2 polymorphism, we use the notion
of let-polymorphism and a `hole', as follows.  A partial evaluator (or
a compiler or interpreter) is an evaluation context that (let-)binds
variables named "app", "lam", etc.  For the object language to
take advantage of let-bound polymorphism in the metalanguage, it
is crucial that we encode object "let" to meta "let".

@@

This strategy of course limits what kinds of processing on encoded
object expressions can be expressed in the object language.  For
example, unlike if we had simply defined |"E"| to be the `identity
encoding' E, we \emph{can} write a size-measurer |SZ[]|, which like |SI[]| is
an object expression with a hole |[]| waiting to be plugged with |"E"|.
However, this notion of size cannot distinguish |let x = E1 in E2[x]|
from |"E2[E1]"|.  Cf. one way to type-check polymorphic let:
\begin{code}
        E1:T    E2[E1]:T'
    ------------------------
    let x = E1 in E2[x] : T'
\end{code}
That analogy shows that the self-interpreter with a hole is a
euphemism for a higher-rank abstraction. That makes it clear where
the difficulty lies, and how to simulate the higher-rank application
(that is, higher-rank arrow elimination) `by hand'.

@@

I guess the let-and-hole argument works out if we don't typecheck the
encoded terms. We can't typecheck the interpreter with the whole
either. We insert the encoded term into the interpreter -- and then
typecheck the whole thing. That'll work. Yet a bit unsatisfactory...

We achieve
self-interpretation in that the following evaluation context in the
object language defines an interpreter for encoded object terms.
\begin{code}
  let lam = \f -> f in
  let app = \f x -> f x in
  let add = \m n -> m + n in
  let int = \i -> i in
  [hole]
\end{code}
The obvious notion of optimality in this setup is easy to achieve.  For
example, given the context above, our partial evaluator turns
\begin{code}
  app (lam (\x -> add x (int 1))) (int 3)
\end{code}
into the number 4, as desired.  We can also encode the self-interpreter
above as the following evaluation context in the object language.
\begin{code}
  let lam = lam (\f -> f) in
  let app = lam (\f -> lam (\x -> app f x)) in
  let add = lam (\m -> lam (\n -> add m n)) in
  let int = lam (\i -> i) in
  [hole]
\end{code}
The "let" and the hole "[hole]" are meta-constructions, which must map to
themselves in any interpretation.

The 'result' of applying an interpreter to itself, needs to be an 
interpreter too!  In fact, if it is Jones-optimal, then you should be 
able to apply it to itself again and get the same answer.  
Our code does that. First, |twice_inc_3_e| is the
metalanguage encoding of the object term |si["e"]|
where si is the self-interpreter of the object language (actually, only
lam,app,add,int), |""| means the object-to-object encoding, and e is the
object term
\begin{code}
  let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
\end{code}

Second, |twice_inc_3_ee| is the metalanguage encoding of the object term
|si["si[e]"]|
which is the same as
|si["si"["e"]]|
The partial evaluator yields 5 on both |twice_inc_3_e| and 
|twice_inc_3_ee|,
as desired.

``But how 
to you create, in either Haskell or MetaOCaml, an untypechecked 
interpreter-with-a-hole [UIH] ?''
Well, one can make an argument that we already have such an
interpreter with polymorphic let and the hole: incope. In Haskell, the
declaration of an instance of Symantics is like the sequence of
polymorphic lets. We construct terms where lam, add, etc, are free
variables. We apply the interpreter to the semantics (plug the hole)
by instantiating these terms (binding the free variables lam, etc. to
the particular instance of Symantics). The unRR construction does
this plugging in explicitly.
Again, what is different from the above is that we \emph{can}
typecheck terms separately, without inserting them first within the
hole of a particular interpreter. Rank-2 type of |repr| helps. It lets
enough of the type information out so the typechecking can
proceed. So, |repr| is the representation of the polymorphic
interpreter context with the hole, which permits separately
typecheckable terms (the evaluation still entails `duplication' so to
speak -- which is one way how polymorphism is resolved).

@@

Boxes go Bananas by Weirich and Washburn.

>>>>

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the following example.
\begin{code}
module EXS(S: Symantics) = struct
 open S
 let diverg () = 
   app (lam (fun x -> int 1)) 
       (app (fix (fun self -> self)) (int 2))
end
\end{code}

If we interpret |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL} realized in OCaml
\begin{code}
let module M = EXS(R) in M.diverg ();;
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = 
    fun k -> k (fun e -> e.ko 
         (fun v -> f {ko = fun k -> k v}))}
end;;
\end{code}
Now when we apply an abstraction to an argument |e|, we always force
the evaluation of that argument before proceeding. This approach is in
line with \cite{ReynoldsXXX} (albeit in a typed setting). The interpreter RCV
is useful for achieving CBV evaluation of the object language
no matter whether the metalanguage is CBV or CBN.

In this section however we show a more general approach to CBV CPS:
a CPS transformer. We take any implementation of |Symantics| (interpreter,
compiler, partial evaluator) and make it evaluate the CPS language
instead. We can view interpreter transformation as a
transformation on the object language, from direct to
continuation-passing style.

The transformation is canonical, as the following (abbreviated) code
shows
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
  let if_ ec et ef = S.lam (fun k ->
    S.app ec (S.lam (fun vc ->
    S.if_ vc (fun () -> S.app (et ()) k) (fun () -> S.app (ef ())
    k))))
  ...
  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
    f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
  let fix f  = S.fix (fun self -> (f self))
end
\end{code}
The code literally establishes a mapping from CPS interpretations to
(direct) interpretations, the latter performed by 
the base interpreter |S|.

A reader may notice that the above module does not define the type
equation for |repr| and so it does not have the signature Symantics.
The reason for that is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has type
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we need a 
type-discriminating
function similar to the type function |repr_pe| of
\S\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom) type variable.

To save space, we show a simplified and less principled approach that
lets us use the module |CPST| as it is. Because the module does not
have the signature Symantics, we cannot apply the |EX| functor to it.
But we can write the tests nevertheless:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. The sample terms are \emph{exactly}
as we used before. We just copied them over. That copying step is the
price we pay for the simplified treatment. We shall also encounter this
copying when discussing the self-interpreter in the next section. We
will see then that the copying step is not frivolous, but represents
a profound instance of 
``plugging the hole'', which is one of the many faces of polymorphism.

@@

which shows the naive CPS transform of the original code. The result is
quite unoptimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with the |P| evaluator instead, our
running test gives the expected
\begin{code}
  P.repr = {P.st = Some <fun>; P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We can easily modify the CBV CPS transform to pass a piece of state
along with the continuation. This lets us support mutable state. We
can also use reference variables of the metalanguage to emulate
reference variables in the object language.
\oleg{Should we have an example in the code?}
\jacques{In the code, yes.  Probably not in the paper, it is getting 
long as it is}

@@

A self-interpreter |si| is an
object-language term such that, for any object-language term e,

@@

where |"e"| is an encoding of |e|.  In our approach, we want to encode
each object-language term as a function from a record%
\footnote{Thanks to currying, the object language need not actually support
records.}  of interpreter methods to a generated representation.  
In the case of a self-interpreter, that
record is simply |si|!  So we should define a self-interpreter |si| to be an
object-language term such that, for any object-language term |e|,

@@

A self-interpreter is then just like the interpreter
|R| in Section 2, except written at the object rather than the meta level.

It is wortwhile noting that, in our class |Symantics|, |add|
is not a term since it does not have type |repr t| for any |t|. We could
have introduced |add| to be of type |repr (Int->Int->Int)| -- but we
didn't. Thus |add| is a special form, which is to be applied in the
meta-language using the meta-language application syntax (i.e. white
space) rather than in the object language (i.e. via |app|).

oleg{Mention the RR interpreter: for any term E, (RR E) is equivalent to E.
RR is not an identity: it is an interpreter that encapsulates another
interpreter. The RR interpreter can be made self if we treat RR as a
special form and interpret it always with itself (similar to let and
hole below).}

Self-interpretation in our framework is somewhat
tricky due to the need for more polymorphism than is easily available
in the meta-language.  An encoded term should
be polymorphic in the interpreter type-constructor "r".  Furthermore, the
functions "lam", "app", etc. should be polymorphic in the object
types "a" and "b".  Thus the object language seems to require rank-2
polymorphism; and so type annotations will be inevitable.

To neatly bypass the need for rank-2 polymorphism, we use the notion
of let-polymorphism and a `hole', as follows.  A partial evaluator (or
a compiler or interpreter) is an evaluation context that (let-)binds
variables named "app", "lam", etc.  For the object language to
take advantage of let-bound polymorphism in the metalanguage, it
is crucial that we encode the object-level "let" using the meta "let".

@@

This strategy of course limits what kinds of processing on encoded
object expressions can be expressed in the object language.  For
example, unlike if we had simply defined |"E"| to be the \emph{identity
encoding} E, we \emph{can} write a size-measure interpreter |SZ[]|, 
which like |SI[]| is
an object expression with a hole |[]| waiting to be plugged with 
an |"E"|.
However, this notion of size cannot distinguish |let x = E1 in E2[x]|
from |"E2[E1]"|.  Recall that one way to type-check polymorphic let is:
\begin{code}
        E1:T    E2[E1]:T'
    ------------------------
    let x = E1 in E2[x] : T'
\end{code}
That analogy shows that the self-interpreter with a hole is a
euphemism for a higher-rank abstraction. That makes it clear where
the difficulty lies, and how to simulate the higher-rank application
(that is, higher-rank arrow elimination) `by hand'.

@@

\oleg{I guess the let-and-hole argument works out if we don't typecheck the
encoded terms. We can't typecheck the interpreter with the hole
either. We insert the encoded term into the interpreter -- and then
typecheck the whole thing. That'll work. Yet a bit unsatisfactory...}

We achieve
self-interpretation in that the following evaluation context in the
object language defines an interpreter for encoded object terms.
\begin{code}
  let lam = \f -> f in
  let app = \f x -> f x in
  let add = \m n -> m + n in
  let int = \i -> i in
  [hole]
\end{code}
The obvious notion of optimality in this setup is easy to achieve.  For
example, given the context above, our partial evaluator turns
\begin{code}
  app (lam (\x -> add x (int 1))) (int 3)
\end{code}
into the number 4, as desired.  We can also encode the self-interpreter
as the following evaluation context in the object language.
\begin{code}
  let lam = lam (\f -> f) in
  let app = lam (\f -> lam (\x -> app f x)) in
  let add = lam (\m -> lam (\n -> add m n)) in
  let int = lam (\i -> i) in
  [hole]
\end{code}
The "let" and the hole "[hole]" are meta-constructions, which must map to
themselves in any interpretation.

The 'result' of applying an interpreter to itself needs to be an 
interpreter too!  In fact, if it is Jones-optimal, then you should be 
able to apply it to itself again and get the same answer.  
Our code does that. First, let |twice_inc_3_e| be the
metalanguage encoding of the object term |si["e"]|,
where si is the self-interpreter for the needed components of the object
language (only |lam|, |app|, |add|, and |int| in this case), |""| means the
object-to-object encoding, and |e| is the object term
\begin{code}
  let twice_ f x = f (f x) in let inc_ n = n + 1 in twice_ inc_ 3
\end{code}
Second, |twice_inc_3_ee| is the metalanguage encoding of the object term
|si["si[e]"]|
which is the same as
|si["si"["e"]]|.
The partial evaluator yields 5 on both |twice_inc_3_e| and 
|twice_inc_3_ee|,
as desired.

\jacques{But how 
to you create, in either Haskell or MetaOCaml, an untypechecked 
interpreter-with-a-hole [UIH] ?}
\oleg{Well, one can make an argument that we already have such an
interpreter with polymorphic let and the hole: incope. In Haskell, the
declaration of an instance of Symantics is like the sequence of
polymorphic lets. We construct terms where lam, add, etc, are free
variables. We apply the interpreter to the semantics (plug the hole)
by instantiating these terms (binding the free variables lam, etc. to
the particular instance of Symantics). The unRR construction does
this plugging in explicitly.}
Again, what is different from the above is that we can
typecheck terms separately, without inserting them first within the
hole of a particular interpreter. Rank-2 type of |repr| helps. It lets
enough of the type information out so the typechecking can
proceed. So, |repr| is the representation of the polymorphic
interpreter context with the hole, which permits separately
typecheckable terms (the evaluation still entails `duplication' so to
speak -- which is one way how polymorphism is resolved).

@@
